---
title: "solubility_clustering"
author: "Alicia Key"
date: "1/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(readr)
library(gridExtra)
theme_set(theme_void())
```

In my prior solubility clustering study, I used hierarchcial clustering to  

Table 1 is a sample of some of the compounds in the dataset:

``` {r}
df <- as_tibble(read.csv("data/delaney-processed.csv")) %>%
  select(
    compound = Compound.ID, 
    mw = Molecular.Weight, 
    h_bond_donors = Number.of.H.Bond.Donors, 
    rings = Number.of.Rings, 
    rotatable_bonds = Number.of.Rotatable.Bonds, 
    psa = Polar.Surface.Area, 
    solubility = measured.log.solubility.in.mols.per.litre
)
knitr::kable(head(df))
```

### Summary of optimum number of clusters

| Solubility vs... | Optimal cluster count |
|---|---|
| polar surface area (PSA) | 3 |
| h-bond donors | 3 |
| rings | 4 |
| rotatable bonds | 3 |
| molecular weight | 4 |

## Linear regressions on each cluster

Using the optimal number of clusters per model, the plots show a linear regression on top of each cluster arrnagement:

``` {r fig.height = 7, fig.width = 7, fig.cap = "Solubility versus property with clusters and linear models"}
alpha <- 0.1
width <- 0.01

hierarchical_clusters <- function(features_df) {
  cluster_labels = list()
  
  distances <- features_df %>%
    as.matrix() %>%
    scale() %>%
    dist()
  
  hclust_out <- hclust(distances, method = "complete")
  
  features_df %>%
    mutate(
      cluster_2 = as.factor(cutree(hclust_out, k = 2)),
      cluster_3 = as.factor(cutree(hclust_out, k = 3)),
      cluster_4 = as.factor(cutree(hclust_out, k = 4))
    )
}

psa_cluster_df <- hierarchical_clusters(select(df, solubility, psa))
h_bond_donor_cluster_df <- hierarchical_clusters(select(df, solubility, h_bond_donors))
rings_cluster_df <- hierarchical_clusters(select(df, solubility, rings))
rotatable_bonds_cluster_df <- hierarchical_clusters(select(df, solubility, rotatable_bonds))
mw_cluster_df <- hierarchical_clusters(select(df, solubility, mw))

psa_cluster_3 <- ggplot(psa_cluster_df, aes(x = psa, y = solubility, col = cluster_3)) +
  geom_jitter(alpha = alpha, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "(a)")

h_bond_donor_cluster_3 <- ggplot(h_bond_donor_cluster_df, aes(x = h_bond_donors, y = solubility, col = cluster_3)) +
  geom_jitter(alpha = alpha, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "(b)")

rings_cluster_4 <- ggplot(rings_cluster_df, aes(x = rings, y = solubility, col = cluster_4)) +
  geom_jitter(alpha = alpha, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "(c)")
  
rotatble_bonds_3 <- ggplot(rotatable_bonds_cluster_df, aes(x = rotatable_bonds, y = solubility, col = cluster_3)) +
  geom_jitter(alpha = alpha, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "(d)")

mw_cluster_4 <- ggplot(mw_cluster_df, aes(x = mw, y = solubility, col = cluster_4)) +
  geom_jitter(alpha = alpha, width = 0.1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "(e)")

grid.arrange(psa_cluster_3, h_bond_donor_cluster_3, rings_cluster_4, rotatble_bonds_3, mw_cluster_4, nrow = 3)
```

## Conclusion

When I use hierarchical clustering to group solubilities, each variable in the dataset needs a different number of clusters to adequately specify its relastionship to solubility.

