---
title: "Solubility Unsupervised"
author: "Alicia Key"
date: "1/23/2021"
output: html_document
---

```{r setup, include=TRUE, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(gridExtra)
library(ranger)
```

## Load the data set

``` {r}
df <- as_tibble(read.csv("data/delaney-processed.csv")) %>%
  select(
    compound = Compound.ID, 
    mw = Molecular.Weight, 
    h_bond_donors = Number.of.H.Bond.Donors, 
    rings = Number.of.Rings, 
    rotatable_bonds = Number.of.Rotatable.Bonds, 
    psa = Polar.Surface.Area, 
    solubility = measured.log.solubility.in.mols.per.litre
)
```

## Extract a matrix

``` {r}
compound_matrix <- df %>%
  select(!c("compound", "solubility")) %>%
  as.matrix()

row.names(compound_matrix) <- df$compound
```

## Run PCA on all the numeric attributes

``` {r}
compound_pca <- prcomp(compound_matrix, scale. = TRUE, center = TRUE)
summary(compound_pca)
```

### Create the plots to explore the PCA

#### biplot

``` {r}
biplot(compound_pca)
```

#### Scree plots

``` {r, fig.caption = "}
pca_pve_df <- tibble(
  principal_component = 1:length(compound_pca_pve),
  proportion_variance_explained = compound_pca_pve,
  cumulative_variance_explained = cumsum(compound_pca_pve)
)

p1 <- ggplot(pca_pve_df, aes(x = principal_component, y = proportion_variance_explained)) +
  geom_line() +
  geom_point(shape = 19, size = 3) +
  ylim(c(0, 1)) +
  labs(x = "principal component",
       y = "proportion of variance explained") +
  theme_minimal()

p2 <- ggplot(pca_pve_df, aes(x = principal_component, y = cumulative_variance_explained)) +
  geom_line() +
  geom_point(shape = 19, size = 3) +
  ylim(c(0, 1)) +
  labs(x = "principal component",
       y = "cumulative proportion of variance explained") +
  theme_minimal()

grid.arrange(p1, p2, nrow = 1)
```

## Hierarchical clustering, based on just molecular weight and solubility

First, run the hierarchical clustering

``` {r}
mw_solubility_matrix <- df %>%
  select(mw, solubility) %>%
  as.matrix() %>%
  scale()
mw_solubility_dist <- dist(mw_solubility_matrix)
mw_solubility_hclust <- hclust(mw_solubility_dist, method = "complete")
plot(mw_solubility_hclust)
```

Now cut a tree for 4 clusters and plot them in a scatter plot coded for their group. Cluster 1 appears to be compounds with high mw and moderate solubility, cluster 2 appears to be compounds with moderate molecular weight and moderate solubility, cluster 3 appeard to be high molecular weight and low solubility, and cluster 4 is low molecular weight and high solubility.

``` {r}
hierarchical_cluster <- as.factor(cutree(mw_solubility_hclust, k = 4))
mw_solubility_clusters <- cbind(df, hierarchical_cluster)

ggplot(mw_solubility_clusters, aes(x = mw, y = solubility, col = hierarchical_cluster)) +
  geom_point(alpha = 0.25) +
  theme_minimal()
```

## Hierarchical clustering with PCA

Use the first principal component. From the biplot above, the first principle component reflects molecular weight, polar surface area, and h bond donors. From looking at the distribution of polarity around a molecule, and how that affects its solubility with water, this makes intuitive sense. From the scree plot above, the first principal component accounts for 75% of the variability in the data, which matches my chemical intuition.

``` {r}
pc1_solubility_matrix <- df %>%
  select(solubility) %>%
  mutate(pc1 = compound_pca$x[, 1]) %>%
  as.matrix() %>%
  scale()
pc1_solubility_dist <- dist(pc1_solubility_matrix)
pc1_solubility_hclust <- hclust(pc1_solubility_dist, method = "complete")
plot(pc1_solubility_hclust)
```

Now plot the clusters within a plot of solubility versus pc1

``` {r}
pc1_hierarchical_cluster <- as.factor(cutree(pc1_solubility_hclust, h = 4))
pc1_solubility_clusters <- df %>%
  transmute(
    solubility,
    compound,
    pc1 = compound_pca$x[, 1],
    pc1_hierarchical_cluster = pc1_hierarchical_cluster
  )

ggplot(pc1_solubility_clusters, aes(x = pc1, y = solubility, col = pc1_hierarchical_cluster)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE)
```

Now let's try both pc1 and pc2, and make a hierarchical cluster between them

``` {r}
pc1_pc2_matrix <- tibble(pc1 = compound_pca$x[, 1], pc2 = compound_pca$x[, 2]) %>%
  as.matrix() %>%
  scale()
pc1_pc2_dist <- dist(pc1_pc2_matrix)
pc1_pc2_hclust <- hclust(pc1_pc2_dist, method = "complete")
plot(pc1_solubility_hclust)
```

Now plot these clusters

``` {r}
pc1_pc2_cluster <- as.factor(cutree(pc1_pc2_hclust, k = 4))
pc1_pc2_clusters <- df %>%
  transmute(
    solubility,
    compound,
    pc1 = compound_pca$x[, 1],
    pc2 = compound_pca$x[, 2],
    pc1_pc2_cluster = pc1_pc2_cluster
  )

pc1_pc2_cluster_mod <- lm(solubility ~ pc1 + pc2 + pc1_pc2_cluster, pc1_pc2_clusters)
summary(pc1_pc2_cluster_mod)
```